---
title: "Class 11: Comparative Structure Analysis"
author: "Grace Wang (PID: A16968688)"
format: html
toc: TRUE
---

## Finishing Class 10

Load the packages we will need for analysis of protein structure sets. 

```{r}
library(bio3d)
#BiocManager::install("msa")
library(msa)
```

We will analyze ADK starting with a single ADK database accession code: "1ake_A". 

```{r}
id <- "1ake_A"
aa <- get.seq(id)
```

Now we can search the PDB database to find all related entries. 

```{r}
blast <- blast.pdb(aa)
```

```{r}
attributes(blast)
head(blast$hit.tbl)
```

Make a summary figure of the results

```{r}
hits <- plot(blast)
```

Our "top hits", i.e., the most similar entries in the database, are :

```{r}
hits$pdb.id
```

```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile = "msa")
```

Sidenote:

```{r}
library(bio3dview)
view.pdbs(pdbs)
```

This is better than directly putting the files into Mol-star, but it is still difficult to see what is similar and different in all these structures or learn about how this family works. 

Let's try PCA. 

```{r}
pc <- pca(pdbs)
plot(pc)
```

```{r}
plot(pc, pc.axes = 1:2)
```

```{r}
view.pca(pc)
```

Write a PDB "trajectory" for Mol-star. 

```{r}
mktrj(pc, file = "pca_results.pdb")
```

