---
title: "Class 12: RNASeq Population Analysis"
author: "Grace Wang (PID: A16968688)"
format: pdf
---

## Question 13

Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding.

```{r}
expr <- read.csv("rs8067378_ENSG00000172057.6.txt", sep = "")

table(expr$geno)
```

```{r}
box <- boxplot(expr$exp ~ expr$geno, 
        pch = 19, 
        xlab = "Genotype", 
        ylab = "Expression")

attributes(box)
box$stats[3, ]
```


Using dplyr to make sure I remember how

```{r message = FALSE, warning=FALSE}
library(dplyr)
```

```{r}
expr_med <- expr %>% group_by(geno) %>% summarize(median = median(exp))
expr_med
```


## Question 14

Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.

```{r}
library(ggplot2)

ggplot(expr) + 
  aes(x = geno, y = exp, fill = geno) + 
  geom_boxplot(notch = T, outliers = F) + 
  geom_point(position = "jitter", alpha = .3)
```

The SNP seems to have an effect on the expression of ORMDL3, as the G/G genotype has lower expression than does the A/A genotype. 